Closure-Free Functional Programming in a Two-Level Type Theory

### 2024-06-18 icfp24paper/paper.tex
fix

Next, we give a logical relation interpretation for $\vW$. We inductively define
a relation in $\whset$ as follows.
\begin{alignat*}{3}
  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
\end{alignat*}
Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of
the function type, which unfolds to pointwise preservation of relations. We also
define $(\vW\,A\,B)\rel\,\gamma\rel\,t\,t'$ to be $\vW\rel\,\gamma\rel\,t\,t'$.
In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to
$\re_\vsup\,\alpha'\,f'$ when the arguments are (inductively) related. Similarly, we
interpret $(\vsup\,t\,u)\rel\,\gamma\rel$ as
$\vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)$. It only remains to
interpret $\vW$-elimination and its $\beta$-rule. We need to define
\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'). \]
We do this by induction on $w\rel\,\gamma\rel :
\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
becomes rather tedious to compute and unfold the involved types, so we only give
a compact definition together with some informal explanation:
\begin{alignat*}{3}
  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
  &\hspace{8.1em}(w\rel\,\gamma\rel)
\end{alignat*}
$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
the induction motive, which matches the goal type. The second argument is the
method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
witnessing the inductive motive for every output of $f\rel$. In the body here,
we need to witness the motive in the case where $w$ is a $\vsup$, which causes
both $\ev\,(\welim\,P\,s\,w)\,\gamma$ and $\re\,(\welim\,P\,s\,w)\,\gamma'$ to
compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
respectively. Hence, we use $s\rel$ to relate the two sides. This definition
also respects $\vW\beta$, by the $\beta$ rule for $\vW\rel$.

>>>>

Next, we give a logical relation interpretation for $\vW$. We inductively define
a relation in $\whset$ as follows.
\begin{alignat*}{3}
  & \vW\rel &&: \Gamma\rel\,\gamma\,\gamma' \to \ev\,(\vW\,A\,B)\,\gamma \to \re\,(\vW\,A\,B)\,\gamma' \to \whset\\
  & \vsup\rel &&: \hspace{0.8em}\{\alpha : \ev\,A\,\gamma\}\{\alpha' : \re\,A\,\gamma'\}(\alpha\rel : A\rel \gamma\rel\,\alpha\,\alpha')\\
  &           &&\to \{f \hspace{0.5em}: \ev\,B\,(\gamma,\,\alpha) \to \ev\,(\vW\,A\,B)\,\gamma\}\\
  &           &&\to \{f' \hspace{0.2em}: \re\,(B \Rightarrow (\vW\,A\,B)[\p])\,(\gamma',\,\alpha')\}\\
  &           &&\to (f\rel : (B \Rightarrow (\vW\,A\,B)[\p])\rel\,(\gamma\rel,\,\alpha\rel)\,f\,f')\\
  &           &&\to \vW\rel\,\gamma\rel\,(\wh{\vsup}\,\alpha\,f)\,(\re_\vsup\,\alpha'\,f')
\end{alignat*}
Here, $(B \Rightarrow (\vW\,A\,B)[\p])\rel$ is the relational interpretation of
the function type, which unfolds to pointwise preservation of relations. We also
define $(\vW\,A\,B)\rel\,\gamma\rel\,t\,t'$ to be $\vW\rel\,\gamma\rel\,t\,t'$.
In short, $\vW\rel$ expresses that $\wh{\vsup}\,\alpha\,f$ is related to
$\re_\vsup\,\alpha'\,f'$ when the arguments are (inductively) related. Similarly, we
interpret $(\vsup\,t\,u)\rel\,\gamma\rel$ as
$\vsup\rel\,(t\rel\,\gamma\rel)\,(u\rel\,\gamma\rel)$. It only remains to
interpret $\vW$-elimination and its $\beta$-rule. We need to define
\[ (\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel : P\rel\,(\gamma\rel,\,w\rel\,\gamma\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'). \]
We do this by induction on $w\rel\,\gamma\rel :
\vW\rel\,\gamma\rel\,(\ev\,w\,\gamma)\,(\re\,w\,\gamma')$. At this point it
becomes rather tedious to compute and unfold the involved types, so we only give
a compact definition together with some informal explanation:
\begin{alignat*}{3}
  &(\vW{-}\msf{elim}\,P\,s\,w)\rel\,\gamma\rel :=\\
  &\hspace{4em}\vW\rel{-}\msf{elim}\,(\lambda\,w\rel.\,\,P\rel\,(\gamma\rel,\,w\rel)\,(\ev\,(\welim\,P\,s\,w)\,\gamma)\,(\re\,(\welim\,P\,s\,w)\,\gamma'))\\
  &\hspace{8.1em}(\lambda\,a\rel\,f\rel\,\mathit{hyp}.\,\,s\rel\,(\gamma\rel,\,a\rel,\,f\rel,\,\mathit{hyp}))\\
  &\hspace{8.1em}(w\rel\,\gamma\rel)
\end{alignat*}
$\vW\rel{-}\msf{elim}$ is the eliminator for $\vW\rel$. Its first argument is
the induction motive, which matches the goal type. The second argument is the
method for $\vsup\rel$; its $a\rel$ and $f\rel$ inputs are the relatedness
witnesses stored in a $\vsup\rel$, while $\mathit{hyp}$ is induction hypothesis,
witnessing the inductive motive for every output of $f\rel$. In the body here,
we need to witness the motive in the case where $w\rel\,\gamma\rel$ is a $\vsup\rel$, which causes
both $\ev\,(\welim\,P\,s\,w)\,\gamma$ and $\re\,(\welim\,P\,s\,w)\,\gamma'$ to
compute further to applications of $\ev\,s\,\gamma$ and $\re\,s\,\gamma'$
respectively. Hence, we use $s\rel$ to relate the two sides. This definition
also respects $\vW\beta$, by the $\beta$ rule for $\vW\rel$.
